Two Problems for Sophistication

### 2015-05-03 Facticity/Facticity/cie-abstract.tex
First attempt

\usepackage{amsmath,amsfonts}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage[mathscr]{eucal}
\usepackage{thmtools}
\usepackage{graphicx}
\usepackage{caption}

\newcommand{\M}{\mathscr M}
\newcommand{\C}{\mathscr C}
\newcommand{\T}{\mathscr T}
\newcommand{\F}{\mathscr F}
\renewcommand{\P}{\mathscr P}
\newcommand{\K}{\mathscr K}
\newcommand{\X}{\mathscr X}
\newcommand{\B}{\mathbb B}
\newcommand{\D}{\Delta}
\newcommand{\N}{\mathbb N}
\newcommand{\tn}[1]{\textnormal{#1}}
\newcommand{\pair}[1]{\left\langle{#1}\right\rangle}
\newcommand{\concat}{\oplus}
\newcommand{\symb}[1]{\texttt{#1}}
\newcommand{\br}[1]{\overline{#1}}
\newcommand{\s}{S}
\newcommand{\dom}[1]{\mathop{\tn{dom}(#1)}}
\newcommand{\range}[1]{\mathop{\tn{range}(#1)}}

\newtheorem{conj}{Conjecture}

\let\doendproof\endproof
\renewcommand\endproof{~\hfill\qed\doendproof}

\newcommand{\p}{\,\text{.}}

\newcommand{\tuple}[1]{\left\langle{#1}\right\rangle}

\newcommand{\hide}[1]{}
\newcommand{\old}[1]{}

\newcommand{\sdr}[1]{\textcolor{blue}{\small #1\textsuperscript{[Steven]} }}
\newcommand{\pb}[1]{\textcolor{OliveGreen}{\small #1 \textsuperscript{[Peter]} }}

\newcommand{\argmin}{\mathop{\arg\min}}

% --- DELETE BEFORE SUBMISSIONS ---
% \pagestyle{headings}

@@

\begin{abstract}
Kolmogorov complexity is a sound, unambiguous formalization of the information content of an object. However, researchers have long been plagued by its inability to express a more intuitive notion of complexity: in terms of Kolmogorov complexity, a completely random bitstring is the most complex object possible. This is not what we mean when we we use the word complexity informally: a television broadcast is complex when we can see structures, faces and landscapes, not when we can only see white noise.

\hspace{0.05\textwidth} The desire to capture this notion of complexity in formal terms has led to many proposals for additional measures. The largest family of proposals is built on the following idea:  an object consists of \emph{structural} and \emph{stochastic} information, and the information content of the structural information determines the complexity of the object. We will use the name \emph{sophistication} for such proposals.

\hspace{0.05\textwidth} In essence, sophistication performs model selection using the class of all computable models. This means that we are dealing not just with the problem of formally defining complexity, but also with a limiting case of statistics: what happens if we extend our model class to the entirety of conceivable models? Do we end up with an ultimate statistics machine, or does the whole framework collapse, and does model selection only work with carefully constrained model classes?

\hspace{0.05\textwidth} While sophistication is a simple enough idea, any attempt to formalize it quickly runs into problems. We highlight the two major issues, and show how they invalidate each proposal. We are forced to conclude that no current proposal truly solves the problem of sophistication, and we offer some informal arguments which suggest that no such formalization may be possible at all.

\end{abstract}

>>>>

@@

\begin{abstract}
Kolmogorov complexity formalizes the idea of information content, but clashes with intuitive ideas of complexity: data seems most complex to us when it contains structure, like language or interesting visual information, not when it contains only random noise.

\hspace{0.05\textwidth}A popular approach is to assume that an object consists of \emph{structural} and \emph{stochastic} information, and that only the former determines the complexity. A dataset $x$ is separated into structural and stochastic parts by representing it as a two-part code: a computable function $\phi$, the \emph{model}, and a string $y$, the \emph{input}, such that $\phi(y) = x$. The aim is that for efficient representations, $\phi$ contains all structure, and $y$ contains all stochastic information. Many proposals, including the structure function, the minimal sufficient algorithmic statistic and sophistication fit this mold. We use \emph{sophistication} as a catch-all term.

\hspace{0.05\textwidth} We highlight two issues. First, it is crucial to correctly measure the information content of the model. All treatments rely on a \emph{numbering}; a canonical description of the computable functions. Kolmogorov complexity is not affected by the choice of numbering by more than a constant, but for sophistication the impact can be dramatic when the length of the plain description of $\phi$ is used to measure information content.

\hspace{0.05\textwidth} Second, we discuss the problems of over- and underfitting. To give an extreme example of underfitting: we can use a universal Turing machine as a model. This representation has all information in the input, yet is maximally efficient, up to a constant. We show that for some numberings, such models are always preferred, leading to a bounded sophistication.

\hspace{0.05\textwidth} Underfitting is a known issue which many treatments seek to circumvent by restricting the class of models to total functions. This leads to the opposite problem: for some numberings, all information ends up inside the model. Only one formalization escapes this problem by allowing non-self-delimiting descriptions. While the question whether this form is invariant remains open, we show that numberings exist that cause it to be bounded for all but extremely deep strings. Thus, it certainly does not fit the original idea behind sophistication.

\hspace{0.05\textwidth} We draw three conclusions. First, sophistication is an interesting problem that leads to diverse questions about statistics, computability and compression. Second, we show that all currently published approaches are either incorrect, or do not behave as desired. Finally we propose that a sound definition of sophistication, capturing the right intuition may be impossible. We ask what we can truly discover about individual objects if our only assumption is that the model should be computable.
\end{abstract}

### 2015-05-03 Facticity/Facticity/paper.tex
First attempt

\newcommand{\M}{\mathscr M}
\newcommand{\C}{\mathscr C}
\newcommand{\T}{\mathscr T}
\newcommand{\F}{\mathscr F}
\renewcommand{\P}{\mathscr P}
\newcommand{\K}{\mathscr K}
\newcommand{\X}{\mathscr X}
\newcommand{\B}{\mathbb B}
\newcommand{\D}{\Delta}
\newcommand{\N}{\mathbb N}
\newcommand{\tn}[1]{\textnormal{#1}}
\newcommand{\pair}[1]{\left\langle{#1}\right\rangle}
\newcommand{\concat}{\oplus}
\newcommand{\symb}[1]{\texttt{#1}}
\newcommand{\br}[1]{\overline{#1}}
\newcommand{\s}{S}
\newcommand{\dom}[1]{\mathop{\tn{dom}(#1)}}
\newcommand{\range}[1]{\mathop{\tn{range}(#1)}}

@@

\begin{abstract}
Kolmogorov complexity measures the amount of information in data, but does not distinguish structure from noise. Kolmogorov's definition of the \emph{structure function} was the first attempt to measure only the structural information in data, by measuring the complexity of an objectively best model under two-part coding. Since then, many variations of this idea have been proposed, for which we use \emph{sophistication} as an umbrella term. We describe two fundamental problems with all existing proposals, showing each  to be \emph{unsound}. Consequently, we put forward the view that the problem is fundamental: it may be impossible to objectively quantify the sophistication.
\end{abstract}

\section{Introduction}\enlargethispage{4ex}
Kolmogorov complexity gives us a sound definition of the amount of information contained in a binary string. It does not, however, capture what most people would consider complexity. For example, a sequence of a million coin flips will almost certainly have maximal Kolmogorov complexity, even though there is nothing complex about flipping a lot of coins---it's just a lot of work. Many scholars have defined alternative complexity measures in the spirit of Kolmogorov complexity, aimed at quantifying not \emph{all} information in a binary string, but only the \emph{meaningful} information in the data: its \emph{sophistication}. In this paper, we investigate two serious problems with sophistication measures, and show why they are unsatisfactory. We believe that the problems are fundamental, and provide arguments why attempts to define sophistication must fail.

The Kolmogorov complexity $C(x)$ of a binary string $x$, is (roughly) the length of the shortest computer program that will output $x$. This length depends on which programming language is used, but, as shown by a central result known as the Invariance Theorem\cite[Section~2.1]{li1993introduction}, the language affects the value only by a constant, independent of $x$. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure of the complexity of binary strings, a \emph{property} of the data. A definition of sophistication $S(x)$ in the spirit of Kolmogorov complexity should have similar robustness guarantees:

\begin{enumerate}
\item $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold: there must be strict limits on how much sophistication can be affected by a change in programming language.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language (except that it's also bounded). Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $C(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
There have been many related proposals for such a measure, that are all based on a \emph{two-part code}: rather than encoding $x$ by means of a single computer program, we encode a \emph{model} for $x$ in the first part of the code, that is interpreted as a representation of $x$'s structural properties. The model does not fully specify $x$, but when combined with the second part of the code, which specifies the noise, the original string becomes fully determined. Thus, the model can be interpreted as a function mapping the noise to the original object.

@@

Underfitting occurs when the selected model is simple, but fails to represent some of the structure present in the data. This is also a problem for sophistication because the considered models are so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such \emph{universal models} are \emph{simple} in the language of sophistication, since they can be described with a relatively small number of bits, yet are able to represent the data within a codelength a constant away from the Kolmogorov complexity. Such a two-part representation essentially encodes all information as noise. If complex models are penalized, then the problem becomes to make sure that universal models are not \emph{always} preferred for complex data. The usual workaround is to restrict the set of allowed models to exclude the universal ones. For example, the model class may be restricted to the set of all total recursive functions. While this excludes universal models, it is questionable whether it adequately solves the problem of underfitting in general.

@@

\paragraph{Related work}
The idea of separating structure from noise originated with Kolmogorov's \emph{structure function} \cite{cover1985kolmogorov}. In this theory, the data $x$ is described as a finite set $A$ that contains $x$, together with the index of $x$ in $A$. The complexity of the set determines the sophistication. This approach was discussed more extensively, and generalised in \cite{vereshchagin2004kolmogorov,gacs2001algorithmic}.

Koppel \cite{koppelSoph1988,koppel1991almost} also built on the structure function work to define a measure of the meaningful information in a binary string, coining \emph{sophistication}, the name we adopt in this paper.

In 1996 Gell-Mann and Lloyd introduced \emph{effective complexity} \cite{gellmann1996information}. Working from the perspective of statistical mechanics, they arrived (independently, it seems) at a notion that fits the mold of sophistication. \emph{Facticity} \cite{adriaans2012facticity}, relaxes the model class to include all partial recursive functions. This publication also identifies the nickname problem discussed in Section~\ref{section:indices}.

Other authors have pointed out problems with sophistication. Antunes et al. \cite{antunes2009sophistication} note that the constant cutoff used by Koppel is a source of instability, a result in the spirit of this paper. They propose a different definition called \emph{coarse sophistication}. This version behaves very differently from regular sophistication, and is closer to computational depth. Vereshchagin \cite{vereshchagin2013algorithmic} also points out that the Kolmogorov minimal sufficient statistic is not always well-defined and suggests requiring that the finite set model should described using a \emph{total} function rather than any computable function. This yields a more robust sophistication that is nonetheless still susceptible to the issues discussed in this paper.
\section{Preliminaries}
Let $\B = \{0,1\}^*$. We deal with partial computable functions $\phi: \B \to \B$, which we will also call \emph{models}. A function is called \emph{prefix} if its domain, $\dom{f} = \{y : f(y) \neq \infty\}$ is a prefix free set, i.e. no string in $\dom{f}$ is a prefix of another. A function $f$ is \emph{total} if $\dom{f} = \B$.

@@

We sometimes need to map binary strings to a prefix-free set. We denote by $\br{x}$ the prefix-encoded representation for $x$. We require that the mapping satisfies $|\br{x}| = |x|+O(\log(|x|)$ (see eg. \cite[Section~1.4]{li1993introduction}). To simplify notation, we will sometimes conflate natural numbers and binary strings, implicitly using the ordering $(0, \epsilon)$, $(1, 0)$, $(2, 1)$, $(3, 00)$, $(4, 01)$, \ldots  
We define the $\M$-Kolmogorov complexity of a binary string $x$, with respect to a model class $\M$ and a numbering $\psi$, as follows $C^{\M,\psi}(x)=\min\{\bar\imath y:\psi_i(y)=x,i\in\M\}$. The model classes $\C$ and $\K$ give us the plain and prefix-free Kolmogorov complexity respectively.

\section{Inefficient indices}\enlargethispage{4ex}
\label{section:indices}
In some works, the definition of sophistication is obtained by opening up the definition of Kolmogorov complexity, and measuring the length of the index in that two-part code. We call such a sophistication \emph{index sophistication}. Examples are Koppel and Atlan's works \cite{koppelSoph1988,koppel1991almost},  and several of its variants \cite{antunes2009sophistication,antunes2013sophistication}.

In these works, no constraints are placed on the numbering. This makes these approaches vulnerable to the \emph{nickname problem} \cite{adriaans2012facticity}: the bits required to specify a model strongly depends on the numbering: \emph{any} index sophistication without requirements on the numbering is either bounded, or nowhere near invariant:

@@

There are two ways to use this notion of model complexity for more robust attempts to define sophistication. Confusingly, both approaches are used in the literature.
First, we can measure the complexity of the \emph{model} using $C^{\M,\psi}(\phi_i)$, which is then the first part of a two-part code describing the data. This approach is used in  \cite{cover1985kolmogorov,gacs2001algorithmic,vitanyi2004meaningful,gellmann1996information} and in most of the present paper.

Second, we can stick to using the length of the index as the measure of sophistication, but require that numbering is \emph{faithful}: for all $i$ there is a $j$ such that $\psi_i=\psi_j$ and $|j|\le C^{\C}(\psi_j)+c$, for some constant $c$. This approach is taken is in \cite{adriaans2012facticity}. We prove in the appendix that there actually do exist faithful, acceptable numberings (Lemma~\ref{lemma:faithful-numberings}). However even choosing a faithful index is not quite enough. The Kolmogorov complexity uses representations of the form $\bar\imath y$, where the bar denotes some straightforward prefix encoding to delimit the model description $i$ from its input $y$. If we define a second prefix encoding $\tilde{\imath}$, with $|\bar\imath|-|\tilde\imath|$ unbounded, we can define a second representation $\bar u \tilde \imath y$, at a constant overhead $|\br{u}|$, and gain more than $|\br{u}|$ for sufficiently complex strings. As shown in Lemma~\ref{lemma:prefix-inefficiency} in the appendix, this results in a bounded sophistication. 
This is in fact an issue with \cite{adriaans2012facticity}. Moreover, we conjecture that \emph{no} computable prefix function $\bar\cdot$ is sufficiently efficient to avoid this.

@@

In the previous section, we saw the first glimpse of the delicate balance between the two code components in sophistication. We will investigate this balance, starting with the sophistication $S^{\K,\psi,c}$. In this setting, we know that, on the one hand, there always exist candidate models---those models $i$ with $i \in \K$ and $C^{\{i\}}(x) \leq C^\K(x) + c$---for which al information except a constant amount is stored in the data, and on the other hand, there are also candidate representations for which all information is stored in the model. This tells us that $S^\K$ is \emph{balanced}: there is no inefficiency in the representation forcing information in or out of the model.
The flip side of having a balanced sophistication is that it becomes easy to show invariance, by tweaking the numbering in such a way that models in a specific subset $\M'\subset\M$ of the model class becomes an arbitrary constant number of bits cheaper to represent than all other models, so that a model in $\M'$ always ends up determining the sophistication.

This means that for model classes with a universal model, we can choose our numbering to make $S^\K$ bounded, or equal to the complexity:
\begin{restatable}[Underfitting]{lemma}{underfitting}
Let $\M$ be a model class containing a universal model $\phi_u$, with the property that $\exists c \forall i \in \M : C^{\{u\}}(x) \leq C^{\{i\}}(x) + c \p$. Then, there exist numberings such that $\s^\M$ is bounded.
\end{restatable}
\begin{restatable}[Overfitting]{lemma}{overfitting}
Let $\M \subseteq \K$ be a model class where for every $x\in\X$ there is a singleton model $i\in\M$ with $\phi_i(\epsilon)=x$. Then there is a numbering $\psi$, and a constant $c$, such that for all $x\in\X$ we have $C^\M(x)-S^{\M,\psi}(x)\leq c$.\label{lemma:overfitting}
\end{restatable}
\noindent Thus, in this balanced sophistication, there is no invariance: all information can be seen as structure, or as noise, depending on the numbering. To avoid these issues, existing proposals upset the balance to exclude or penalize at least the singleton models and the universal models.

@@

The existing proposals that can overfit are the minimal sufficient statistic in the structure function \cite{cover1985kolmogorov}, the sophistication as defined in \cite{mota2013sophistication}, the minimal sufficient algorithmic statistic \cite{gacs2001algorithmic} and Vereshchagin's strongly sufficient statistic \cite{vereshchagin2013algorithmic}. It is widely recognized that singleton models are always candidates in these settings, but not, it seems, that they may be the only candidates for any string, as Lemma~\ref{lemma:overfitting} shows.

@@

\subsection{Boosting a submodel by manipulating the numbering}

\overfitting*

Now, by Lemma~\ref{lemma:thecoolone} there is an numbering $\psi_1,\psi_2,\ldots$ such that $\s^{\M,\phi}(x)=|\bar 0|+S^{\M',\psi}(x)$. We observed that $|C^{\K}(f)-C^{\K}(x)|\le c$ for any $f\in\M'$, so $\s^{\M',\psi}(x)\ge C^{\K}(x)-c$. This proves the lemma.
\end{proof}
\subsection{Underfitting}

\underfitting*

\begin{proof}
Let $D$ be a prefix function as in Lemma~\ref{lemma:thecoolone} such that it returns the index of $u$ for the argument $\epsilon$ and $\infty$ for any other argument. That is, $\M' = \{u\}$. This construction satisfies the conditions 1 and 2 from Lemma~\ref{lemma:thecoolone}; invoking it we find that there exists an acceptable numbering $\psi$ for which $\s^{\M,\psi}(x) = \s^{\M', phi}(x) + c$. Since $\M'$ contains only a single model, $\s_{\M',\phi}(x)$ is constant.
\end{proof}

>>>>

\newcommand{\M}{\mathscr M}
\newcommand{\C}{\mathscr C}
\newcommand{\T}{\mathscr T}
\newcommand{\F}{\mathscr F}
\renewcommand{\P}{\mathscr P}
\newcommand{\K}{\mathscr K}
\newcommand{\X}{\mathscr X}
\newcommand{\B}{\mathbb B}
\newcommand{\D}{\Delta}
\newcommand{\N}{\mathbb N}
\newcommand{\tn}[1]{\textnormal{#1}}
\newcommand{\pair}[1]{\left\langle{#1}\right\rangle}
\newcommand{\concat}{\oplus}
\newcommand{\symb}[1]{\texttt{#1}}
\newcommand{\br}[1]{\overline{#1}}
\newcommand{\s}{S}
\newcommand{\bi}{\bar\imath}
\newcommand{\dom}[1]{\mathop{\tn{dom}(#1)}}
\newcommand{\range}[1]{\mathop{\tn{range}(#1)}}

@@

\begin{abstract}
Kolmogorov complexity measures the amount of information in data, but does not distinguish structure from noise. Kolmogorov's definition of the \emph{structure function} was the first attempt to measure only the structural information in data, by measuring the complexity of the smallest model allowing for optimal compression of the data. Since then, many variations of this idea have been proposed, for which we use \emph{sophistication} as an umbrella term. We describe two fundamental problems with all existing proposals, showing each  to be \emph{unsound}. Consequently, we put forward the view that the problem is fundamental: it may be impossible to objectively quantify the sophistication.
\end{abstract}

\section{Introduction}

Kolmogorov complexity gives us a sound definition of the amount of information contained in a binary string. It does not, however, capture what most people would consider complexity. For example, a sequence of a million coin flips will almost certainly have maximal Kolmogorov complexity, even though there is nothing complex about flipping a coin repeatedly. Many scholars have defined additional complexity measures in the spirit of Kolmogorov complexity, aimed at quantifying not \emph{all} information in a binary string, but only the \emph{meaningful} information. While this concept has been given many names, we use \emph{sophistication} as an umbrella term. In this paper, we investigate two serious problems with sophistication measures. We conclude with three arguments that suggest the problems are fundamental, explaining our belief that sophistication cannot be defined in a satisfactory manner.

The Kolmogorov complexity $C(x)$ of a binary string $x$, is (roughly) the length of the shortest computer program that will output $x$. This length depends on which programming language is used, but, as shown by a central result known as the Invariance Theorem \cite[Section~2.1]{li1993introduction}, the language affects the value only by a constant, independent of $x$. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure of the complexity of binary strings, a property \emph{of the data.} A definition of sophistication $S(x)$ in the spirit of Kolmogorov complexity should have similar robustness guarantees:

\begin{enumerate}
\item $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold: there must be strict limits on how much sophistication can be affected by a change in programming language.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language (except that it's also bounded). Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $C(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
There have been many related proposals for such a measure, all based on a \emph{two-part code}: rather than encoding $x$ by means of a single computer program, we encode a \emph{model} for $x$ in the first part of the code, that is interpreted as a representation of $x$'s structural properties. The model does not fully specify $x$, but when combined with the second part of the code, which specifies the noise, the original string becomes fully determined. Thus, the model can be interpreted as a function mapping the noise to the original object.

@@

Underfitting occurs when the selected model is simple, but fails to represent some of the structure present in the data. This is also a problem for sophistication because the models under consideration are so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such \emph{universal models} are \emph{simple} in the language of sophistication, since they can be described with a relatively small number of bits, yet are able to represent the data within a codelength a constant away from the Kolmogorov complexity. Such a two-part representation essentially encodes all information as noise. If complex models are penalized, then the problem becomes to make sure that universal models are not \emph{always} preferred for complex data. The usual workaround is to restrict the set of allowed models to exclude the universal ones. For example, the model class may be restricted to the set of all total recursive functions. While this excludes universal models, it is questionable whether it adequately solves the problem of underfitting in general.

@@

\section{Preliminaries}
One of the main issues in dealing with sophistication in a general sense, is that one must unify many different approaches, all founded on slightly different principles. We have devised a general framework which, while it may not represent the idiosyncracies of all methods, allows us to rephrase them in a uniform way that illustrates the properties we discuss. We hope the reader will trust that these properties remain also in the original statements of the measures.

Let $\B = \{0,1\}^*$. We deal with partial computable functions $\phi: \B \to \B$, which we will also call \emph{models}. A function is called \emph{prefix} if its domain, $\dom{f} = \{y : f(y) \neq \infty\}$ is a prefix free set, i.e. no string in $\dom{f}$ is a prefix of another. A function $f$ is \emph{total} if $\dom{f} = \B$.

@@

We sometimes need to map binary strings to a prefix-free set. We denote by $\br{x}$ the prefix-encoded representation for $x$. We require that the mapping satisfies $|\br{x}| = |x|+O(\log(|x|)$ (see eg. \cite[Section~1.4]{li1993introduction}). To simplify notation, we will sometimes conflate natural numbers and binary strings, implicitly using the ordering $(0, \epsilon)$, $(1, 0)$, $(2, 1)$, $(3, 00)$, $(4, 01)$, \ldots

For technical reasons, we deviate slightly from the traditional notation of Kolmogorov complexity: we define the $\M$-Kolmogorov complexity of a binary string $x$, with respect to a model class $\M$ and a numbering $\psi$, as follows $C^{\M,\psi}(x)=\min\{\bar\imath y:\psi_i(y)=x,i\in\M\}$. This means that $C^\C(x)$ corresponds to the plain Kolmogorov complexity, traditionally denoted $C(x)$, and $C^\K(x)$ corresponds to the prefix-free Kolmogorov complexity $K$. This notation also allows us to represent the smallest possible description of $x$ using a single model $\phi_i$ as $C^{\{i\}}(x)$. Note that this includes the description of $i$ as well as its shortest program for $x$.

Finally we not that the choice of numbering, combined with a choice of prefix function, serves the same purpose that is expressed in the choice of a universal Turing machine. Requiring an acceptable numbering and a computable prefix function is equivalent to choosing a universal Turing machine: this requirement suffices to make the Kolmogorov complexity invariant. As we shall see, however simply choosing a universal Turing machine is not enough for a proper treatment of sophistication: we must choose our numbering carefully. Thus, we phrase our requirement for an invariant sophistication with respect to the \emph{numbering} chosen.    There should be strict guarantees on how much the value of $S(x)$ can changes if the numbering is varied.

\section{Inefficient indices}
\label{section:indices}

The simplest approach to sophistication would simply be to `open up' the Kolmogorov complexity and to see which program on the universal Turing machine achieves the smallest description length: the program that \emph{witnesses} the Kolmogorov complexity. Since any UTM is defined in terms of a two-part coding, such programs necessarily consist of a model and an input. While this approach is not literally used anywhere in the literature, it is an illustrative point to start our investigation. As is common, we will allow all witnesses within a constant slack term $c$ from the Kolmogorov complexity, and choose the one with the smallest model.

\begin{definition}[Index sophistication]
\[
s_{\text{index}}^{\M_0,\M_1,\psi,c}(X) = \min\left\{ |i| : \i in \M_0, C^{\{i\}, \psi}(x) \leq C^{\M_1, \psi}(x)+c   \right\}
\]
\end{definition}

Koppel and Atlan's treatment \cite{koppelSoph1988,koppel1991almost}, where the name \emph{sophistication} originates, follows this basic logic, although it contains many idiosyncracies like the use ot monotonic models, and an inherent extension to infinite strings. As the subsequent history of sophistication has discarded these, we will not discuss them here.

In \cite{antunes2009sophistication} Koppel's principle is first recast to finite strings. The definition used is similar to $s_\text{index}^{\T, \C,\psi,c}$, with one exception: the total complexity of a witness $(i,y)$ is measured as $|i|+|y|$ without the cost of delimiting the two. This differences are not relevant to the point we make in this section. The restriction to total classes is a common technique to avoid underfitting, which we will discuss in the next section.

@@

Note that such perverse numberings are not an issue for Kolmogorov complexity. If our universal Turing machine uses an inefficient numbering, we can switch to a moe efficient one at only a constant cost. This gives us an insight into what happens inside Kolmogorov complexity. We may define the program $\bi y$ using a model $i$ and an input $y$, but if there's any inefficiency in the description of $i$, the shortest program will be achieved by using $i$ to define a more efficient model encoding, and encoding both the model and its input in $y$. For the Kolmogorov complexity, such considerations are academic, since we only look at the total length of the program. For sophistication, however, they are crucial.

There are two ways to use the model complexity $C^\M(\phi_i)$ for more robust attempts to define sophistication. Confusingly, both are used in the literature. First, we can measure the complexity of the \emph{model} using $C^{\M,\psi}(\phi_i)$, which is then the first part of a two-part code describing the data. This approach is used in  \cite{cover1985kolmogorov,gacs2001algorithmic,vitanyi2004meaningful,gellmann1996information} and in most of the present paper.

Second, we can stick to using the length of the index as the measure of sophistication, but restrict the allowed numberings. This approach is taken is in \cite{adriaans2012facticity}. Adriaans  defines \emph{facticity} as $s_\text{index}^{\C,\C,\psi,0}$, but only allows numberings that are \emph{faithful}: for all $i$ there is a $j$ such that $\psi_i=\psi_j$ and $|j|\le C^{\C}(\psi_j)+c$, for some constant $c$.

We prove in the appendix that---contrary to Adriaans' suggestion---there actually do exist faithful, acceptable numberings (Lemma~\ref{lemma:faithful-numberings}). However even choosing a faithful index is not quite enough. The Kolmogorov complexity uses representations of the form $\bar\imath y$, where the bar denotes some straightforward prefix encoding to delimit the model description $i$ from its input $y$. If we define a second prefix encoding $\tilde{\imath}$, with $|\bar\imath|-|\tilde\imath|$ unbounded, we can define a second representation $\bar u \tilde \imath y$, at a constant overhead $|\br{u}|$, and gain more than $|\br{u}|$ for sufficiently complex strings. As shown in Lemma~\ref{lemma:prefix-inefficiency} in the appendix, this results in a bounded sophistication (thus the facticity is bounded).

@@

In the previous section, we saw the first glimpse of the delicate balance between the two code components in sophistication. We will investigate this balance, starting with the variant $S^{\K,\psi,c}$. For reasons which will become clear, this variant is not used anywhere in the literature, but it helps to illustrate the issues we wich to discuss.

In this setting, we know that, on the one hand, there always exist witnesses for which all information except a constant amount is stored in the input, and on the other hand, there are also candidate representations for which all information is stored in the model. This tells us that $S^\K$ is \emph{balanced}: there is no inefficiency in the representation forcing information in or out of the model.
The flip side of having a balanced sophistication is that it becomes easy to show a lack of invariance, by tweaking the numbering in such a way that models in a specific subset $\M'\subset\M$ of the model class becomes an arbitrary constant number of bits cheaper to represent than all other models, so that a model in $\M'$ always ends up determining the sophistication.

This means that for model classes with a universal model, we can choose our numbering to make $S^\K$ bounded, or equal to the complexity:

\begin{restatable}[Underfitting]{lemma}{underfitting}
Let $\M$ be a model class containing a universal model $\phi_u$, with the property that $\exists c \forall i \in \M : C^{\{u\}}(x) \leq C^{\{i\}}(x) + c \p$. Then, there exist numberings such that $\s^\M$ is bounded.
\end{restatable}

For this reason, many treatments restrict the model class. What is perhaps less well known is that the same property holds in the other direction: there are numberings for which the singleton models always determine the sophistication.

\begin{restatable}[Overfitting]{lemma}{overfitting}
Let $\M \subseteq \K$ be a model class where for every $x\in\X$ there is a singleton model $i\in\M$ with $\phi_i(\epsilon)=x$. Then there is a numbering $\psi$, and a constant $c$, such that for all $x\in\X$ we have $C^\M(x)-S^{\M,\psi}(x)\leq c$.\label{lemma:overfitting}
\end{restatable}

The proof of both these lemmas relies on a simple principle: for certain subclasses $\M' \subset \M$, there exist numberings which, when used, have the effect of penalizing the codelength $C^\K(\phi_i)$ for any model outside $\M'$ by an arbitrary constant amount. We can use this penatly to effectively `push' these models outside of the range of considered models, ensuring that, under this numbering, a model in $\M'$ always determines the sophistication.

For the lemmas above this means setting $\M' = {u}$, with $\phi_u$ a universal model, or letting $\M'$ be the set of all singletons. The requirements for $\M'$ are strict, and slightly complex. The following lemma illustrates the principle.

Lemmas~\ref{lemma:underfitting} and \ref{lemma:overfitting} follow as corollaries. For Lemma~\ref{lemma:underfitting}:
\begin{proof}
Let $D$ be a prefix function as in Lemma~\ref{lemma:thecoolone} such that it returns the index of $u$ for the argument $\epsilon$ and $\infty$ for any other argument. That is, $\M' = \{u\}$. This construction satisfies the conditions 1 and 2 from Lemma~\ref{lemma:thecoolone}; invoking it we find that there exists an acceptable numbering $\psi$ for which $\s^{\M,\psi}(x) = \s^{\M', phi}(x) + c$. Since $\M'$ contains only a single model, $\s_{\M',\phi}(x)$ is constant.
\end{proof}

And for Lemma~\ref{lemma:overfitting}

Now, by Lemma~\ref{lemma:thecoolone} there is an numbering $\psi_1,\psi_2,\ldots$ such that $\s^{\M,\phi}(x)=|\bar 0|+S^{\M',\psi}(x)$. We observed that $|C^{\K}(f)-C^{\K}(x)|\le c$ for any $f\in\M'$, so $\s^{\M',\psi}(x)\ge C^{\K}(x)-c$. This proves the lemma.
\end{proof}

\noindent Thus, in this balanced sophistication, there is no invariance: all information can be seen as structure, or as noise, depending on the numbering. To avoid these issues, existing proposals upset the balance to exclude or penalize at least the singleton models and the universal models.

@@

We will now review the treatments in the literature that show overfitting. The first is the structure function, proposed by Kolmogorov himself. Most likely the first attempt at separating structure from noise in an object manner. Kolmogorov defined the function
\[
h_x(\alpha) = \min \left \{\log |S| : S \ni x, C^\K(s) \leq \alpha \right \} 
\]
and suggested that the smallest set for which $K(S) + \log|S| \leq C^\K(S) + c$ for some pre-chosen constant$c$ can be seen as capturing all the structural information in $x$. \footnote{\pb{Discuss how much of this Kolmogorov believed himself and note that this was unpublished work.}} This is equivalent to the sophistication $\s^{\F,\K,\psi,c}$. Lemma~\ref{lemma:underfitting} shows us that given $c$, there are numberings for which this sophistication is always equal, up to a constant, to $C^\K(x)$. Thus, while the general shape of the structure function might tell an interesting story, its sophistication function does not satisfy our requirements.

$\s^{\F,\K,\psi,c}$ is also used in \cite{antunes2013sophistication}. \pb{This version doesn't work at all\ldots}

Also the regular and naive versions from \cite{mota2013sophistication} \ldots

the structure function \cite{cover1985kolmogorov}

the sophistication as defined in \cite{mota2013sophistication} (and the naive soph?)

the minimal sufficient algorithmic statistic \cite{gacs2001algorithmic}

Vereshchagin's strongly sufficient statistic \cite{vereshchagin2013algorithmic}. \pb{fix citation}

It is widely recognized that singleton models are always candidates in these settings, but not, it seems, that they may be the only candidates for any string, as Lemma~\ref{lemma:overfitting} shows.

@@
